---
title: "FAC-eda"
author: "C.Robertson"
date: "2023-10-16"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

## Data Description

The data set includes the following data fields:

- `Departure Terminal` - 
- `Arrival Terminal` -
- `Vessel` - 
- `Scheduled Depature Time` - 
- `Actual Departure Time` -  
- `Bus`, `Semi`, `Comm`, `Pvoh`, `Pvuh`, `Mtcy` - Bus, Commercial, Passenger Overheight, Passenger Underheight, Motorcyle
- `Total Veh` - Total Vehicles
- `Foot Pass` - Number of Foot Passengers
- `Veh Pass` -  Number of Passengers in Vehicles
- `Total Pass` - Total Number of Passengers
- `Total O/L` - Total overloaded (unclear if this is an estimate based on # left in the lot..)

The time period covered by this data set is from Nov 2021 to April 2023, with the transition to the two ferry system using *Kwigis* and *Gwawis* ferries occurring on April 12, 2022. After some investigation, it appears the passenger breakdown (Foot vs Vehicle) is not consistently populated, but Totals of passengers should be accurate.


```{r echo = FALSE, warning=FALSE, message=FALSE}
library("readr")
library("readxl")
library("ggplot2")
library("dplyr")
library("lubridate")
df3 <- read_excel("data/Rte 19 - Sail by Sail Nov 22 - Apr 23.xlsx", 
     col_types = c("text", "text", "text", 
         "text", "numeric", "date", "date", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric"))
df2 <- read_excel("data/Rte 19- Sail by Sail May 22 - Oct 22.xlsx", 
     col_types = c("text", "text", "text", 
         "text", "numeric", "date", "date", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric"))
df1 <- read_excel("data/Rte 19 - Sail by Sail Nov 21 - Apr 22 .xlsx", 
     col_types = c("text", "text", "text", 
         "text", "numeric", "date", "date", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric"))
df <- rbind(df1, df3) #df2, df3)
df$Hour <- as.numeric(format(df$`Sched Dept Ts`, '%H'))
df$Year <- as.numeric(format(df$`Sched Dept Ts`, '%Y'))
df$Month <- as.numeric(format(df$`Sched Dept Ts`, '%m'))
df$Day <- as.numeric(format(df$`Sched Dept Ts`, '%d'))
```



## Observed Patterns

### General Traffic Pattern - Comparing Nov-Apr for Quinsam (2021-2022) and Island Class (2022-2023))

```{r echo = FALSE, warning=FALSE, message=FALSE}
df <- df %>% mutate(DayOfWeek = wday(`Sched Dept Ts`, label = TRUE))


df$Ferry <- df$Vessel
df$Ferry[df$Ferry %in% c("Island Gwawis", "Island Kwigwis")] <- "Island Class"

summary_data <- df %>%
  group_by(Hour, DayOfWeek, Ferry) %>%
  summarize(TotalVehicles = sum(`Total Veh`),
            TotalPassengers = sum(`Total Pass`)) %>% filter(Ferry %in% c("Quinsam", "Island Class"))

# Total Vehicles by Time of Day as Faceted Bar Graph, split by Departure Terminal
ggplot(summary_data, aes(x = Hour, y = TotalVehicles, fill = Ferry)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~DayOfWeek, ncol = 2) +
  labs(title = "Total Vehicles by Time of Day, Nov 21 - Apr 22 (Quinsam), Nov 22 - Apr 23 (Island Class)",
       x = "Hour of Day",
       y = "Total Vehicles")  +
  theme_minimal() + theme(legend.position="bottom")

# Total Passengers by Time of Day as Faceted Bar Graph, split by Departure Terminal
# Total Vehicles by Time of Day as Faceted Bar Graph, split by Departure Terminal

ggplot(summary_data, aes(x = Hour, y = TotalPassengers, fill = Ferry)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~DayOfWeek, ncol = 2) +
  labs(title = "Total Passengers by Time of Day, Nov 21 - Apr 22 (Quinsam), Nov 22 - Apr 23 (Island Class)",
       x = "Hour of Day",
       y = "Total Passengers") +
  theme_minimal() + theme(legend.position="bottom")

df2 <- df %>% mutate(WeekdayOrWeekend = ifelse(DayOfWeek %in% c("Sat", "Sun"), "Weekend", "Weekday")) %>%
  mutate(Month =format(`Actual Dept Ts`, "%b"))


# Create summary data
summary_data_over_time <- df2 %>%
  group_by(Month, Ferry) %>% filter(Ferry %in% c("Quinsam", "Island Class")) %>%
  summarize(SumTotalVehicles = sum(`Total Veh`),
            SumTotalPassengers = sum(`Total Pass`))

ggplot(filter(summary_data_over_time, Month != "Apr"), aes(x = factor(Month, levels=c("Nov", "Dec", "Jan", "Feb", "Mar", "Apr")), y = SumTotalVehicles, group=Ferry)) +
  geom_line(aes(color=Ferry)) + geom_point() +
  labs(title = "Total Vehicles Over Time (by Month)",
       x = "Month",
       y = "Total Vehicles")  +
  theme_minimal() + theme(legend.position="bottom")

ggplot(filter(summary_data_over_time, Month != "Apr"), aes(x = factor(Month, levels=c("Nov", "Dec", "Jan", "Feb", "Mar", "Apr")), y = SumTotalPassengers, group=Ferry)) +
  geom_line(aes(color=Ferry)) + geom_point() +
  labs(title = "Total Passengers Over Time (by Month)",
       x = "Month",
       y = "Total Passengers") +
  theme_minimal() + theme(legend.position="bottom")

```

## Totals over each period

```{r echo = FALSE, warning=FALSE, message=FALSE}
summary_data_over_time = summary_data_over_time %>% mutate(passengerperVehicle=SumTotalPassengers/SumTotalVehicles)
sum_totals = summary_data_over_time %>% group_by(Ferry) %>% summarise(SumTotalVehicles = sum(SumTotalVehicles), SumTotalPassengers = sum(SumTotalPassengers), passengerperVehicle=mean(SumTotalPassengers/SumTotalVehicles))
sum_total = sum_totals %>% mutate(passengerperVehicle=SumTotalPassengers/SumTotalVehicles)
knitr::kable(summary_data_over_time)
knitr::kable(sum_total)
```


## Island Class Ferries, May 2022 - Apr 2023


```{r echo = FALSE, warning=FALSE, message=FALSE}

df3 <- read_excel("data/Rte 19 - Sail by Sail Nov 22 - Apr 23.xlsx", 
     col_types = c("text", "text", "text", 
         "text", "numeric", "date", "date", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric"))
df2 <- read_excel("data/Rte 19- Sail by Sail May 22 - Oct 22.xlsx", 
     col_types = c("text", "text", "text", 
         "text", "numeric", "date", "date", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric", "numeric", "numeric", 
         "numeric"))
df <- rbind(df2, df3)
df$Hour <- as.numeric(format(df$`Sched Dept Ts`, '%H'))
df$Year <- as.numeric(format(df$`Sched Dept Ts`, '%Y'))
df$Month <- as.numeric(format(df$`Sched Dept Ts`, '%m'))
df$Day <- as.numeric(format(df$`Sched Dept Ts`, '%d'))
df$Delay <- df$`Actual Dept Ts` - df$`Sched Dept Ts`
df$Delay <- as.numeric(df$Delay) / 60
df <- df %>% mutate(DayOfWeek = wday(`Sched Dept Ts`, label = TRUE))

summary_data <- df %>%
  group_by(Hour, DayOfWeek, `Departure Terminal`) %>%
  summarize(TotalDelay = sum(Delay),
            TotalOverLoad = sum(`Total O/L`))
summary_data2 <- df %>%
  group_by(Hour, Month, `Departure Terminal`) %>%
  summarize(TotalDelay = sum(Delay),
            TotalOverLoad = sum(`Total O/L`))
summary_data2$Month <- factor(summary_data2$Month, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) 
summary_data2 <- summary_data2 %>% filter(TotalDelay>=0)
```

### Overloads

```{r echo = FALSE, warning=FALSE, message=FALSE}
ggplot(summary_data, aes(x = Hour, y = TotalOverLoad, fill=`Departure Terminal`)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~DayOfWeek, ncol = 2) +
    labs(title = "Overloads by Time of Day, May 22 - Apr 23",
         x = "Hour of Day",
         y = "Total # of Overloads") +
    theme_minimal()  + theme(legend.position="bottom")

ggplot(summary_data2, aes(x = Hour, y = TotalOverLoad, fill=`Departure Terminal`)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~Month, ncol = 2) +
    labs(title = "Overloads by Time of Day, May 22 - Apr 23",
         x = "Hour of Day",
         y = "Total # of Overloads") +
    theme_minimal()  + theme(legend.position="bottom")

```


### Delays

```{r echo = FALSE, warning=FALSE, message=FALSE}

ggplot(summary_data, aes(x = Hour, y = TotalDelay/60, fill=`Departure Terminal`)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~DayOfWeek, ncol = 2) +
    labs(title = "Delay (Hrs) by Time of Day, May 22 - Apr 23",
         x = "Hour of Day",
         y = "Total Delay Time (Hrs)") +
    theme_minimal()  + theme(legend.position="bottom")

ggplot(summary_data2, aes(x = Hour, y = TotalDelay/60, fill=`Departure Terminal`)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~Month, ncol = 2) +
    labs(title = "Delay (Hrs) by Time of Day, May 22 - Apr 23",
         x = "Hour of Day",
         y = "Total Delay Time (Hrs)") +
    theme_minimal()  + theme(legend.position="bottom")
    
```

### Efficiency

Comparing # of vehicles to # of passengers (totals).

```{r echo = FALSE, warning=FALSE, message=FALSE}
df <- df %>% dplyr::filter(`Total Veh` > 0)
df <- df %>% mutate(Efficiency = `Total Pass` / (`Total Veh`))

ggplot(df, aes(x=`Total Veh`, y = `Total Pass`)) + geom_point(aes(color=`Departure Terminal`)) +
    labs(title = "Passenger Total vs. Vechicle Total",
         x = "Total Vehicle Count",
         y = "Total Passenger Count") + theme_minimal() + theme(legend.position="bottom")

summary_data <- df %>%
  group_by(Hour, DayOfWeek, `Departure Terminal`) %>%
  summarize(Efficiency = mean(Efficiency))

ggplot(summary_data, aes(x = Hour, y = Efficiency, fill = `Departure Terminal`))  +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~DayOfWeek, ncol = 3) +
  labs(title = "Efficiency by Time of Day",
       x = "Hour of Day",
       y = "Efficiency") +
  theme_minimal()
#2. What travel patterns emerge from existing data?  From observation, it is clear there is a pattern to #the commercial traffic.  What about the largest demographic on our island, older citizens in smaller #households?  When do they travel?  Yes, we believe this is part of the congestion on a weekly basis on #Thursdays, but when else are those “experience card holders” travelling?  How, if needed, could this #demand be relocated.

#3. Has there been an increase in the frequency of travel basis upon the schedule (more frequent sailings) #from a particular set of “experience card holders” and when?  In other words, have our existing travellers #changed their travel habits?

```

